Active Learning from Blackbox to Timed Connectors

### 2016-01-13 paper/paper.tex
.

% personal characters
\newcommand{\rblock}[0]{\circleddash}
\newcommand{\rread}[0]{\smiley}
\newcommand{\rnoread}[0]{\frownie}

@@

@@

In this paper, we proposed a active learning framework (shown in Figure \ref{fig:howto}) to
automatically extract timed connectors from blackbox models.

As a branch of \emph{machine learning} technique, active learning directly applies queries to a
low-level blackbox model and construct hypothesis according to the results.
There're already several works like \cite{DBLP:journals/mt/Daelemans10,
DBLP:journals/iandc/Angluin87, DBLP:conf/fase/RaffeltS06} where active learning is used to extract
\emph{finite state automatas} or \emph{grammars}.

We present \emph{Parameterized Mealy machine} provide a parametierized semantics for timed Reo
channels, which enables us to generate Mealy machines with a given alphabet. Then the $L*$ algorithm
is adapted and optimized to handle Mealy machines with time action.

The paper is structured as follows. After this general introduction, in Section
\ref{sec:preliminaries} we briefly illustrate some basic concepts, including Reo the coordination
language, Mealy machine, and active automata learning. Section \ref{sec:semantics} defines the
Mealy-machine-based operational semantics of Reo, which is used in Section \ref{sec:activelearning}
to show how to extract Reo models from blackboxes by means of active learning. Finally, in Section
\ref{sec:experiment}, we discuss the optimization and implementation in \texttt{Golang}.

\section{Preliminar ies} 
\label{sec:preliminaries}
\subsection{Reo Coordination Language} 
\label{sec:reo}
We provide here a brief overview of the main concepts of Reo, more details can be found in
\cite{DBLP:journals/mscs/Arbab04, DBLP:journals/scp/BaierSAR06}.

@@

\begin{definition}[Mealy machine]
  A Mealy machine is a 6-tuple $(S, s_0, I, O, T, G)$ consisting of the following:
  \begin{itemize}
    \item[-] a finite set of states $S$
    \item[-] a start state (also called initial state) $s_0$ which is an element of $S$
    \item[-] a finite set called the input alphabet $I$
    \item[-] a finite set called the output alphabet $O$
    \item[-] a transition function $\delta : S \times I \rightarrow S$ mapping pairs of a
      state and an input symbol to the corresponding next state.
    \item[-] an output function $\lambda : S \times I \rightarrow O$ mapping pairs
      of a state and an input symbol to the corresponding output symbol.
  \end{itemize}
\end{definition}

@@

\section{Timed Connectors as Mealy Machines}
\label{sec:semantics}
\subsection{External Behavior of Reo Connectors}
So far as we can tell, most
works\cite{DBLP:conf/fase/RaffeltS06, DBLP:journals/corr/ChenHLLTWW15} on automata learning are not
capable of infinite models. \todo{reason?} Considering the semantics defined in section
\ref{sec:reo}, it's apparent that every finite connector has a corresponding constraint
automata, and the automata is also finite. Unfortunately, when checking the further definition of
Reo connectors' input, we find that it's not the case.

\todo{an example as graph}

\begin{figure}[h]
  \caption{Infinite States in a Reo Model}
  \label{fig:reoinfinite}
\end{figure}

While focusing on behavior of connectors, Reo doesn't give detail depiction on behavior of
components. As shown in \figurename \ref{fig:reoinfinite}, connectors are able to reject any datum
if they are not ready. But what will happen outside the connector, if the datum is rejected? This
question deserves careful consideration if we are taking an external view.

\subsection{Time Domain}
Time is involved in several extension version of Reo. For example, Timed
Reo\cite{DBLP:conf/sefm/ArbabBBR04}, Hybrid Reo\cite{DBLP:conf/icfem/ChenSS14}, etc.
Generally, these models are designed to handle real-time behavior where time is defined in
$\mathbb{R}$. Besides, we also found some works\cite{DBLP:journals/fmsd/PrabhakarDM015} rational
time is chosen to make things easier. In this paper, we choose the rational number field
$\mathbb{Q}$ as our time domain, which simplifies discretization of timed behaviors greatly.

@@

\subsection{Parameterized Mealy Machine}
We present a model named \emph{parameterized mealy machine} to represent timed connectors
(hereinafter referred to as PMM).

This model is supposed to behave as a middle representation. Connectors are defined as parameterized
mealy machines, and composed via its production operator. Then original mealy-machine model will be
taken as semantics of PMM model.

\begin{definition}[Parameterized Mealy Machine]
  A \emph{Parameterized Mealy Machine} is defined as a function $\mathcal{PM}(\Sigma)=\langle
  S(\Sigma), s_0, I, O, \delta(\Sigma), \lambda(\Sigma)\rangle$ that maps an
  alphabet to its corresponding Mealy Machine. 
  \begin{itemize}
    \item[-] $\Sigma$ is a \emph{finite} datum alphabet (hereinafter referred to as an alphabet)
    \item[-] $S$ is a function that maps an alphabet to a \emph{finite} set of
      states. We use $S(\Sigma)$ to denote the state set.
    \item[-] $I$ is a finite set of source-ends.
    \item[-] $O$ is a finite set of sink-ends.
    \item[-] $s_0$ is the initial state. It satisfies $\forall \Sigma,s_0\in S(\Sigma)$
    \item[-] $\delta$ maps a \emph{finite} datum alphabet to an \emph{output function}. We use
      $\delta(\Sigma):S(\Sigma)\times Input(\Sigma,I)\rightarrow Output(\Sigma, O)$ to denote the output function.
    \item[-] $\lambda$ maps an alphabet to a \emph{transition function}. We use
      $\lambda(\Sigma):S(\Sigma)\times Input(\Sigma,I)\rightarrow S(\Sigma)$ to denote the transiton
      function.
  \end{itemize}
\end{definition}

In the definition above, \emph{Input} and \emph{Output} are used to generate input actions and output actions by the corresponding alphabets and ends.
\todo{How to say?}
\[
Input(\Sigma,I)=\{\}\cup\{T\}
\]
where we use $T$ to denote the time action.
\[
Output(\Sigma,I)=\{\}
\]

Now we can use Parameterized Mealy Machines to define a new semantics to Reo.
\begin{example}[PMM Semantics of FIFO channel $PM_{FIFO}$]
  \label{example:pmmfifo}
  The semantics of FIFO channel with source end $A$ and sink end $B$ can be defined as follows.
  \begin{itemize}
    \item[-] $S(\Sigma)=\{q_0\}\cup\{q_d|d\in\Sigma\}$
    \item[-] $I=\{A\}$
    \item[-] $O=\{B\}$
    \item[-] $s_0=q_0$
    \item[-] output function
      \begin{displaymath}
        \delta(\Sigma)(s,i)=\left\{
        \begin{array}[h]{ll}
          (B:\bot) & s=q_0\land i=(A:\_,B:\rnoread) \\
          (B:d) & s=q_0\land i=(A:d,B:\rread)\\
          \rblock & s=q_0\land i=(A:\bot,B:\rread) \\
          (B:d) & s=q_d\land i=(A:\_,B:\rread)\\
          \rblock & s=q_d\land i=(A:d,B:\rnoread) \\
          (B:\bot) & s=q_d\land i=(A:\_,B:\rnoread) \\     
        \end{array}
        \right.
      \end{displaymath}
    \item[-] transition function
      \begin{displaymath}
        \lambda(\Sigma)(s,i)=\left\{
        \begin{array}[h]{ll}
          q_d & s=q_0\land i=(A:d,B:\rnoread) \\
          q_d & s=q_{d'}\land i=(A:d,B:\rread) \\
          q_0 & otherwise \\
        \end{array}
        \right.
      \end{displaymath}
  \end{itemize}
\end{example}

Besides, the concrete mealy machine (where $\Sigma=\{a\}$) is shown in Figure \ref{fig:pmmfifo}.
\begin{figure}[h]
  \begin{center}
    \input{./images/pmmfifo.tex}
  \end{center}
  \caption{PMM-based Semantics of FIFO, where $\Sigma=\{a\}$}
  \label{fig:pmmfifo}
\end{figure}

@@

\section{From Blackbox to Timed Connectors} 
\label{sec:activelearning}

@@

\subsection{Case Studies}

@@

\section{Conclusion and Future Work}

>>>>

% personal characters
\newcommand{\rblock}[0]{\circleddash}
\newcommand{\rread}[0]{\rhd}
\newcommand{\rnoread}[0]{\oslash}
\newcommand{\smap}[1]{[{#1}]}

@@

As a branch of \emph{machine learning} technique, active learning offers a way to obtain models from
low-level blackboxes. Works in \cite{DBLP:journals/mt/Daelemans10, DBLP:journals/iandc/Angluin87,
DBLP:conf/fase/RaffeltS06} shows impressive examples where active learning is used to extract
\emph{finite state automatas} or \emph{grammars}.

\todo{better expressions!}

@@

In this paper, we proposed a active learning framework (shown in Figure \ref{fig:howto}) to
automatically extract timed connectors from blackbox models. We present \emph{Parameterized Mealy
machine} as a parametierized semantics for timed Reo channels, which enables us to generate Mealy
machines with a given alphabet. Then the $L*$ algorithm is adapted and optimized to extract Mealy
machines with time action from the \emph{connectors in blackboxes}.

The rest of the paper is organized as follows. After this general introduction, in Section
\ref{sec:preliminaries} we briefly illustrate some basic concepts, including Reo the coordination
language, Mealy machine, and active automata learning. Section \ref{sec:semantics} defines the
Mealy-machine-based operational semantics of Reo, which is used in Section \ref{sec:activelearning}
to show how to extract Reo models from blackboxes by means of active learning. Finally, in Section
\ref{sec:experiment}, we discuss the optimization and implementation in \texttt{Golang}.

\section{Preliminaries} 
\label{sec:preliminaries}
\subsection{Reo Coordination Language} 
\label{sec:reo}
We provide here a brief overview of the main concepts of Reo, more details can be found in
\cite{DBLP:journals/mscs/Arbab04, DBLP:journals/scp/BaierSAR06}.

@@

\begin{definition}[Mealy machine]
  A Mealy machine is a 6-tuple $(S, s_0, I, O, \delta, \lambda)$ consisting of the following:
  \begin{itemize}
    \item[-] a finite set of states $S$
    \item[-] a start state (also called initial state) $s_0$ which is an element of $S$
    \item[-] a finite set called the input alphabet $I$
    \item[-] a finite set called the output alphabet $O$
    \item[-] a transition function $\delta : S \times I \rightarrow S$ mapping pairs of a
      state and an input symbol to the corresponding next state.
    \item[-] an output function $\lambda : S \times I \rightarrow O$ mapping pairs
      of a state and an input symbol to the corresponding output symbol.
  \end{itemize}
\end{definition}

@@

\section{Timed Connectors as Mealy Machines}
\label{sec:semantics}
In this section, we show how a timed connector be transformed into a Mealy machine with time action.

Since time is not involed in original Mealy machine, we first discuss how to formalize the time
dimension in Mealy machine and timed Connectors. After that, we present a \emph{parameterized Mealy
machine} as a bridge between connectors and Mealy machines.

\subsection{Time Domain}
Time is involved in several extension version of Reo. For example, Timed
Reo\cite{DBLP:conf/sefm/ArbabBBR04}, Hybrid Reo\cite{DBLP:conf/icfem/ChenSS14}, etc.
Generally, these models are designed to handle real-time behavior where time is defined in
$\mathbb{R}$. Besides, we also found some works like \cite{DBLP:journals/fmsd/PrabhakarDM015} where
rational time indeed  makes things easier. In this paper, we choose the rational number field
$\mathbb{Q}$ as our time domain, which simplifies discretization of timed behaviors greatly.

@@

\subsection{Parameterized Mealy Machine}
We present a model named \emph{parameterized Mealy machine} (hereinafter referred to as PMM)
to represent timed connectors. PMM is supposed to behave as a middle representation. Connectors are
defined as parameterized mealy machines, and composed via its production operator. Then original
mealy-machine model will be taken as semantics of PMM model.

Following the formal definition of Mealy machine in Section \ref{sec:preliminaries}, we present the
definition of \emph{parameterized Mealy machine} as follows.

\begin{definition}[Parameterized Mealy Machine]
  A \emph{Parameterized Mealy Machine} is defined as a 6-tuple $\mathcal{PM}=\langle
  S(\Sigma), s_0, I, O, \delta(\Sigma), \lambda(\Sigma)\rangle$ where 
  \begin{itemize}
    \item[-] $\Sigma$ is a \emph{finite} datum alphabet (hereinafter referred to as an alphabet)
    \item[-] $S$ is a function that maps an alphabet to a \emph{finite} set of
      states. We use $S(\Sigma)$ to denote the state set.
    \item[-] $I$ is a finite set of source-ends.
    \item[-] $O$ is a finite set of sink-ends.
    \item[-] $s_0$ is the initial state. It satisfies $\forall \Sigma,s_0\in S(\Sigma)$
    \item[-] $\delta$ maps a \emph{finite} datum alphabet to an \emph{output function}. We use
      $\delta(\Sigma):S(\Sigma)\times Input(\Sigma,I)\rightarrow Output(\Sigma, O)$ to denote the output function.
    \item[-] $\lambda$ maps an alphabet to a \emph{transition function}. We use
      $\lambda(\Sigma):S(\Sigma)\times Input(\Sigma,I)\rightarrow S(\Sigma)$ to denote the transiton
      function.
  \end{itemize}
\end{definition}

In the definition above, \emph{Input} and \emph{Output} are used to generate a set of input actions
and output actions from the corresponding alphabets and ends. For example, if we have a alphabet
consisting of two items $d_0$ and one source-end named $A$, the input actions would be
\begin{eqnarray*}
  Input(\{d_0\},\{A\}) &=& \{\{A:d_0,B:\rread\},\{A:d_0,B:\rnoread\}, \\
  & & \{A:\bot,B:\rread\},\{A:\bot,B:\rnoread\},T\}
\end{eqnarray*}
where we use $\rread$ to indicate that a end-sink is ready for write, and $\rnoread$ otherwise.

\todo{How to say?}
\[
Input(\Sigma,I)=\{\}\cup\{T\}
\]
where we use $T$ to denote the time action.
\[
Output(\Sigma,I)=\{\}
\]

Parameterized Mealy machines can be seen as an abstract form of Mealy machines, which means that we
still need a convert function between the two models.

\begin{definition}[Concretize Mapping]
  We use $\smap{\mathcal{PM}}_{\Sigma}$ to denote the concrete Mealy machine which is determined by
  an abstract parameterized Mealy machine $\mathcal{PM}$ and the alphabet $\Sigma$. Apparently
  $\smap{\mathcal{PM}}_{\Sigma}$ can be described as
  \begin{eqnarray*}
    \smap{\mathcal{PM}}_{\Sigma} &=& 
    \langle
    \mathcal{PM}.S(\Sigma), \mathcal{PM}.s_0, \\
    & & Input(\Sigma, \mathcal{PM}.I), Output(\Sigma, \mathcal{PM}.O), \\
    & & \mathcal{PM}.\delta(\Sigma), \mathcal{PM}.\lambda(\Sigma),
    \rangle
  \end{eqnarray*}
\end{definition}

Now we can use Parameterized Mealy Machines to define a new semantics for timed Reo. Here we
take one asynchronous channel (FIFO1), one asynchronous channel (Sync) and one timed channel (Timer)
as examples.

\begin{example}[PMM Semantics of FIFO1 channel]
  \label{example:pmmfifo}
  The semantics of FIFO channel with source end $A$ and sink end $B$ can be defined as follows.
  \begin{itemize}
    \item[-] $S(\Sigma)=\{q_0\}\cup\{q_d|d\in\Sigma\}$
    \item[-] $I=\{A\}$, $O=\{B\}$, $s_0=q_0$
    \item[-] output function
      \begin{displaymath}
        \delta(\Sigma)(s,i)=\left\{
        \begin{array}[h]{ll}
          (B:\bot) & s=q_0\land i=(A:\_,B:\rnoread) \\
          (B:d) & s=q_0\land i=(A:d,B:\rread)\\
          \rblock & s=q_0\land i=(A:\bot,B:\rread) \\
          (B:d) & s=q_d\land i=(A:\_,B:\rread)\\
          \rblock & s=q_d\land i=(A:d,B:\rnoread) \\
          (B:\bot) & s=q_d\land i=(A:\_,B:\rnoread) \\     
        \end{array}
        \right.
      \end{displaymath}
    \item[-] transition function
      \begin{displaymath}
        \lambda(\Sigma)(s,i)=\left\{
        \begin{array}[h]{ll}
          q_d & s=q_0\land i=(A:d,B:\rnoread) \\
          q_d & s=q_{d'}\land i=(A:d,B:\rread) \\
          q_0 & otherwise \\
        \end{array}
        \right.
      \end{displaymath}
  \end{itemize}
\end{example}

Besides, the concrete mealy machine (where $\Sigma=\{a\}$) is shown in Figure \ref{fig:pmmfifo}.
\begin{figure}[h]
  \begin{center}
    \input{./images/pmmfifo.tex}
  \end{center}
  \caption{PMM-based Semantics of $\smap{FIFO1}_\Sigma$, where $\Sigma=\{a\}$}
  \label{fig:pmmfifo}
\end{figure}

\begin{example}[PMM Semantics of Sync channel]
  The PMM-based semantics of a Sync channel with a source-end A and a sink-end B can be defined as:
  \begin{itemize}
    \item[-] $S(\Sigma)=\{q_0\}$, $I=\{A\}$, $O=\{B\}$, $s_0=q_0$
    \item[-] output function
      \begin{displaymath}
        \delta(\Sigma)(s,i)=\left\{
        \begin{array}[h]{ll}
          (B:\bot) & s=q_0\land i=(A:\bot,B:\rnoread) \\
          (B:d) & s=q_0\land i=(A:d,B:\rread)\\
          \rblock & s=q_0\land i=(A:\bot,B:\rread) \\
          \rblock & s=q_0\land i=(A:d,B:\rnoread) \\
        \end{array}
        \right.
      \end{displaymath}
    \item[-] transition function $\lambda(\Sigma)(s,i)=q_0$.
  \end{itemize}
\end{example}

\begin{example}[PMM Semantics of n-Timer channel]
  Considering a n-Timer channel with a source-end A and a sink-end B, we define its PMM-based
  semantics as:
  \begin{itemize}
    \item[-] $S(\Sigma)=\{q_0\}\cup\{q_d|d\in\Sigma\}$
    \item[-] $I=\{A\}$, $O=\{B\}$, $s_0=q_0$
    \item[-] output function
      \begin{displaymath}
        \delta(\Sigma)(s,i)=\left\{
        \begin{array}[h]{ll}
          (B:\bot) & s=q_0\land i=(A:\_,B:\rnoread) \\
          (B:d) & s=q_0\land i=(A:d,B:\rread)\\
          \rblock & s=q_0\land i=(A:\bot,B:\rread) \\
          (B:d) & s=q_d\land i=(A:\_,B:\rread)\\
          \rblock & s=q_d\land i=(A:d,B:\rnoread) \\
          (B:\bot) & s=q_d\land i=(A:\_,B:\rnoread) \\     
        \end{array}
        \right.
      \end{displaymath}
    \item[-] transition function
      \begin{displaymath}
        \lambda(\Sigma)(s,i)=\left\{
        \begin{array}[h]{ll}
          q_d & s=q_0\land i=(A:d,B:\rnoread) \\
          q_d & s=q_{d'}\land i=(A:d,B:\rread) \\
          q_0 & otherwise \\
        \end{array}
        \right.
      \end{displaymath}
  \end{itemize}
\end{example}

@@

\section{From Blackbox to Timed Connectors} 
\label{sec:activelearning}
In this section, we will show how to use L* algorithm to extract models from blackboxes. Firstly, we
presents the definition of \emph{Observation Table}, which is used to describe the hypothesis.

@@

\subsection{Case Studies}
A simple example of timed connector is presented to show how L* works.
\begin{figure}[h]
  \begin{center}
    \input{./images/expfifo.tex}
  \end{center}
  \caption{Expiring FIFO1 Channel}
  \label{fig:expfifo}
\end{figure}

@@

\section{Conclusion and Future Work}
Future work should focus on better representation on timed connectors and implementation of
equivanlence query.
